Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-02-09 numerics_paper/version2/numerical_chj.tex

As we will see, carbuncles also appear in the numerical solution of the shallow
water circular hydraulic jump.  This is natural, since the solution involves a
standing shock wave.  Dealing with the carbuncle in this context is particularly
interesting and challenging, since this standing shock should (at least in an
appropriate flow regime) be unstable, and some research has suggested that the
carbuncle is the manifestation of a true physical
instability \cite{moschetta2001carbuncle,elling2009carbuncle}.

\begin{itemize}
  \item Describe the carbuncle instability starting with Euler and going to SWEs.
  \item Mention that we observe the cc instability for the CHJ as well.
  \item Describe that in this case, the situation is more complicated than in other problems since we
    expect the correct solution to have physical instabilities triggered by round off error.
  \item Based on existing literature we know that we can use highly dissipative and robust methods to
    avoid the formation of carbuncle instabilities. However, in this case, these methods also tend to
    dissipate other important features of the solution unless the mesh is highly refined.
  \item We are interested in methods that prevent the formation of cc instability in the CHJ
    but preserve other features and physical instabilities.
  \item There are multiple ideas and methods proposed to fix the cc insitability for the Euler
    and recently for the SWEs.
  \item It is believed that the cc instability occurs due to lack or reduced viscosity in the shear contact waves.
  \item Some methods also propose to improve the entropy stability properties of the method. In particular, to induce or guarantee that the entropy is dissipated at shocks.
  \item Based on some of these ideas we start with Rusanov's method using the cell averages of the solution.
    This is a first-order method that is entropy stable and positivity preserving. We enhance its robustness by
    overestimating the wave speed when strong shear is present in the solution. The result is a robust solver
    that do not develop cc instabilities. However, the method is highly dissipative and, as a result,
    many other important features of the solution are dissipated.
  \item We then blend this highly dissipative method with a FV method based on Roe's average.
  \item After this we introduce the second order corrections from Law-Wendroff's method.
\end{itemize}

\subsection{Our contribution}
\begin{itemize}
\item We present a new problem where the cc instability occurs. In our opinion, this is a more challenging
  problem than the standard bow shock problem since we expect the solution to be non-steady.
  We aim to preserve instabilities that are inherently present in the CHJ for certain regimes. And,
  at the same time, we aim to prevent the cc instability.
\item We explore the use of entropy dissipation and entropy stability as cure of the cc instability for the SWEs.
\item We propose a specific method to blend the highly dissipative and robust Rusanov's RP and the
  accurate 3-wave Roe's solver for the SWEs.
\item We demonstrate that this scheme prevents the formation of cc instabilities not only
  in the well-known bow shock problem, but also in the CHJ while preserving other important features of
  the solution.
\end{itemize}

\section{The shallow water circular hydraulic jump}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
%The system \eqref{eq:sw} can be written in vector form as
%\begin{align}
%    q_t + f(q)_x + g(q)_y & = 0.
%\end{align}

\subsection{Semi-analytical steady solution under rotational symmetry}\label{sec:steady_chj}
In this work, we consider the initial boundary value problem consisting of the
shallow water model \eqref{eq:sw} in an annular domain
\begin{align}
r_\textup{jet} \le r \le r_\infty
\end{align}
where $r = \sqrt{x^2+y^2}$, with prescribed inflow at $r=r_\textup{jet}$
and prescribed outflow at $r=r_\infty$.
The domain and boundary conditions are rotationally symmetric.
By assuming rotational symmetry in \eqref{eq:sw}, one obtains the system
\begin{subequations} \label{eq:rsw}
\begin{align}
    (rh)_t + (rhu)_r & = 0 \label{mass1} \\
    (rhu)_t + (rhu^2)_r + r \left(\frac{1}{2}gh^2\right)_r = 0. \label{mom1}
\end{align}
\end{subequations}
where the depth $h$ and radial velocity $u$ are functions of the single spatial coordinate $r$.
Steady-state solutions of \eqref{eq:rsw} satisfy
\begin{subequations}\label{steady}
\begin{align}
    rhu & = C \\
    h'(r) & = \frac{h}{\frac{g}{\beta^2} r^3 h^3 -r} = \frac{h}{r} \cdot \frac{F^2}{1-F^2} \label{eq:dh}
\end{align}
\end{subequations}
for some $C$ independent of $r$.  Here $F=|u|/\sqrt{gh}$ is the Froude number.
We see that two types of steady profiles exist, depending on whether the flow
is subcritical ($|F|<1$) or supercritical ($|F|>1$).  No smooth steady solution can
include both regimes, since the right hand side of \eqref{eq:dh} blows up when $F=1$.

\section{Numerical methods}
In this work we study the behavior of certain shock-capturing finite volume
methods based on the use of Riemann solvers.  For simplicity, we discuss these
solvers in the context of a 1-dimensional mesh.  
In the numerical experiments in \S\ref{sec:num} we use second-order Strang
splitting \cite{strang1968construction} to extend the method to multiple dimensions.

\subsection{Wave propagation methods}\label{sec:waveprop}
The numerical solution $q_i$ represents the average value of the set of
conserved quantities over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(q^n_{i-1},q^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the LWL scheme
\begin{align}\label{second-order_via_fluct}
  q_i^{n+1} = q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, \\
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align}\label{correction-fluxes}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a TVD-limited version of $\W_{i\pm 1/2}^p$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  q_i^{n+1} = q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
\end{align}
which, for the conservative Riemann solvers we use herein, can also be written
as a first-order flux-differencing scheme:
\begin{align}\label{first-order_FV}
  q_i^{n+1}=q_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}$ is a numerical approximation of the flux at $x_\iph$.

@@

As we discussed in the introduction, finite volume methods that use
a Roe Riemann solver are prone to the carbuncle instability for the CHJ.
In contrast, methods that use a Rusanov Riemann solver exhibit no carbuncles. 
However, such methods tend to dissipate other important physical features of
the solution, unless the grid is highly refined.
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle-free but does not dissipate important features of the solution.
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers.
Our blending criterion is based on the residual of the entropy (in)equality,
which induces dissipation of entropy near discontinuities.
Blending these Riemann solvers has been proposed before
to solve the Euler equations (see
\cite{nishikawa2008very,wang2016developing,jaisankar2007diffusion,ohwada2018simple,deng2019new,ray2013entropy}),
the shallow water equations (see \cite{bader2014carbuncle,kemm2014note}),
and even the Navier-Stokes equations (see \cite{nishikawa2008very,ohwada2018simple}).
Our approach differs from those just cited in that it is based on the local entropy;
for a related approach in the context of the Euler equations, see
\cite{ismail2009affordable,ismail2009proposed}.

@@

>>>>

As we will see, carbuncles can appear in the numerical solution of the shallow
water circular hydraulic jump.  This is natural, since the solution involves a
standing shock wave.  Dealing with the carbuncle in this context is particularly
interesting and challenging, since this standing shock should (at least in an
appropriate flow regime) be unstable, and some research has suggested that the
carbuncle is the manifestation of a true physical
instability \cite{moschetta2001carbuncle,elling2009carbuncle}.

\subsection{Outline}
In this work, we describe and study the circular hydraulic jump as a new
test problem for shallow water discretizations.  We find that although some
existing methods behave acceptably on previous test problems, they
are not capable of providing accurate solutions for the circular hydraulic jump
across the range of flow regimes we study.  We develop a new entropy-stable
Riemann solver for the shallow water equations that generates solutions with
the correct behavior.

\section{Numerical methods for the shallow water equations}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + f(q)_x + g(q)_y & = 0.
\end{align}

In this work we study the behavior of certain shock-capturing finite volume
methods based on the use of Riemann solvers.  For simplicity, we discuss these
solvers in the context of a 1-dimensional problem and mesh.  
In the numerical experiments in \S\ref{sec:num} we use second-order Strang
splitting \cite{strang1968construction} to extend the method to multiple dimensions.

\subsection{Wave propagation methods}\label{sec:waveprop}
Let $Q_i$ represent the average value of the set of
conserved quantities over cell $i$, which extends from $x_\imh$ to $x_\iph$.
At each time step and at each interface $x_\imh$, we approximately solve the
Riemann problem with initial states $(Q^n_{i-1},Q^n_i)$.  The approximate
solution consists of a set of traveling discontinuities or waves $\W^p_\imh$
and corresponding speeds $s^p_\imh$.
We use the wave propagation framework developed by LeVeque \cite{leveque1997wave, leveque2002finite}
and employed in the Clawpack software \cite{clawpack,pyclaw-sisc}, which implements the 
Lax-Wendroff-LeVeque (LWL) scheme
\begin{align}\label{second-order_via_fluct}
  Q_i^{n+1} = Q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$,
the fluctuations are given by
\begin{align}\label{fluct}
  \apdq_\imh := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \amdq_\iph := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p, \\
\end{align}
and represent the effect (to first
order accuracy) of waves coming from the solution of the Riemann problem at
each of the neighboring interfaces.  Meanwhile, $\Ft_{i\pm 1/2}$ are correction
fluxes that make the method second-order accurate:
\begin{align}\label{correction-fluxes}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p.
\end{align}
Here $\tilde{\W}_{i\pm 1/2}^p$ is a TVD-limited version of $\W_{i\pm 1/2}^p$.
We will sometimes work with
the first-order method obtained by setting the correction fluxes to zero:
\begin{align}\label{first-order_via_fluct}
  Q_i^{n+1} = Q_i^n-\frac{\Delta t}{\Delta x}\left[\apdq_\imh+\amdq_\iph\right]
\end{align}
which, for the conservative Riemann solvers we use herein, can also be written
as a first-order flux-differencing scheme:
\begin{align}\label{first-order_FV}
  Q_i^{n+1}=Q_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}$ is a numerical approximation of the flux at $x_\iph$.

@@

As we discussed in the introduction, finite volume methods that use a
Roe Riemann solver are prone to exhibit the carbuncle instability.
In contrast, methods that use a Rusanov Riemann solver exhibit no carbuncles. 
However, such methods tend to dissipate other important physical features of
the solution, unless the grid is highly refined.
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle-free but does not dissipate important features of the solution.
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers.
Our blending criterion is based on the residual of the entropy (in)equality,
which induces dissipation of entropy near discontinuities.
Blending these Riemann solvers has been proposed before
to solve the Euler equations (see
\cite{nishikawa2008very,wang2016developing,jaisankar2007diffusion,ohwada2018simple,deng2019new,ray2013entropy}),
the shallow water equations (see \cite{bader2014carbuncle,kemm2014note}),
and even the Navier-Stokes equations (see \cite{nishikawa2008very,ohwada2018simple}).
Our approach differs from those just cited in that it is based on the local entropy;
for a related approach in the context of the Euler equations, see
\cite{ismail2009affordable,ismail2009proposed}.

@@

\section{Numerical study of the circular hydraulic jump}

\subsection{Semi-analytical steady solution under rotational symmetry}\label{sec:steady_chj}
In this section, we consider the initial boundary value problem consisting of the
shallow water model \eqref{eq:sw} in an annular domain
\begin{align}
r_\textup{jet} \le r \le r_\infty
\end{align}
where $r = \sqrt{x^2+y^2}$, with prescribed inflow at $r=r_\textup{jet}$
and prescribed outflow at $r=r_\infty$.
The domain and boundary conditions are rotationally symmetric.
By assuming rotational symmetry in \eqref{eq:sw}, one obtains the system
\begin{subequations} \label{eq:rsw}
\begin{align}
    (rh)_t + (rhu)_r & = 0 \label{mass1} \\
    (rhu)_t + (rhu^2)_r + r \left(\frac{1}{2}gh^2\right)_r = 0. \label{mom1}
\end{align}
\end{subequations}
where the depth $h$ and radial velocity $u$ are functions of the single spatial coordinate $r$.
Steady-state solutions of \eqref{eq:rsw} satisfy
\begin{subequations}\label{steady}
\begin{align}
    rhu & = C \\
    h'(r) & = \frac{h}{\frac{g}{\beta^2} r^3 h^3 -r} = \frac{h}{r} \cdot \frac{F^2}{1-F^2} \label{eq:dh}
\end{align}
\end{subequations}
for some $C$ independent of $r$.  Here $F=|u|/\sqrt{gh}$ is the Froude number.
We see that two types of steady profiles exist, depending on whether the flow
is subcritical ($|F|<1$) or supercritical ($|F|>1$).  No smooth steady solution can
include both regimes, since the right hand side of \eqref{eq:dh} blows up when $F=1$.
